Radio Crickets: Chirping Jets from Black Hole Binaries Entering their Gravitational Wave Inspiral

### 2015-12-09 Paper/paper.tex
Finalise paper for resubmission

\begin{abstract}
  We study a novel electromagnetic signature of supermassive black
  hole binaries whose inspiral starts being dominated by gravitational
  wave (GW) emission.  Recent simulations suggest that the binary's
  member BHs can continue to accrete gas from the circumbinary
  accretion disk in this phase of the binary's evolution, all the way
  until coalescence.  If one of the binary members produces a radio
  jet as a result of accretion, the jet precesses along a biconical
  surface due to the binary's orbital motion.  When the binary enters
  the GW phase of its evolution, the opening angle widens, the jet
  exhibits milliarcsecond scale wiggles, and the conical surface of
  jet precession is twisted due to apparant superluminal motion.  The
  rapidly increasing orbital velocity of the binary gives the jet an
  appearance of a ``chirp.''  This helical chirping morphology of the
  jet can be used to infer the binary parameters.  For binaries with
  mass $10^7$--$10^{10}$ M$_\odot$ at redshifts $z<0.5$, monitoring
  these features in current and archival data will place a lower limit
  on sources that could be detected by eLISA and Pulsar Timing Arrays.
  In the future, microarcsecond interferometry with the Square
  Kilometer Array will increase the potential usefulness of this
  technique.
\end{abstract}

@@

Binaries of supermassive black holes (BHs) arise naturally as a result
of mergers of galaxies in the context of heirarchical galaxy formation
\citep{2000MNRAS.311..576K, 2002MNRAS.336L..61H, 2008ApJ...676...33D,
  2012MNRAS.422.1306K}.  After a galaxy merger, infalling black holes
lose their angular momentum in three stages before coalescence
\citep{1980Natur.287..307B, 2005LRR.....8....8M, 2011ASL.....4..181C}.
In the first stage, the black holes sink to the center of the
gravitational potential of the merger remnant due to dynamical
friction and form a gravitationally bound binary.  The newly formed
binary then continues to lose its angular momentum and shrink by
scattering ambient gas and stars.  In the final stage of the binary's
evolution, emission of gravitational waves (GWs) becomes the
predominant mode of angular momentum loss, which results in the
coalescence of the black holes.

General relativisitic (GR) numerical simulations can predict the
binary's evolution in its final stage up to coalescence
\citep{2005PhRvL..95l1101P, 2006PhRvL..96k1102B, 2006PhRvL..96k1101C}.
This has stimulated interest in predicting related electromagnetic
signals of black hole coalescence.  Effects considered in the
literature include afterglows due to infall of gas on the remnant
black hole \citep{2005ApJ...622L..93M}, electromagnetic variability in
the circumbinary disk due to shocks induced by the sudden mass loss of
the binary following GW emission at coalescence
\citep{2007APS..APR.S1010B, 2008ApJ...676L...5L}, flares from shocked
remnants of accretion disk around the recoiled remnant black hole
\citep{2007PhRvL..99d1103L, 2008ApJ...682..758S, 2009CQGra..26i4032H},
quasi-periodic variability in gas accretion in the early stages of
black hole merger \citep{2008ApJ...672...83M, 2009CQGra..26i4032H,
  2012MNRAS.427.2680K, 2013MNRAS.436.2997D, 2014ApJ...783..134F,
  2015ApJ...807..131S}, evolution in the profiles of broad emission
lines shortly before merger \citep{2013MNRAS.432.1468M},
characteristic evolution in thermal emission due to circumbinary
accretion \citep{2015MNRAS.446L..36F}, brightening of the circumbinary
accretion disk due to viscous dissipation of gravitational waves
\citep{2008PhRvL.101d1101K}, and prompt tidal disruption of ambient
stars by the recoiled remnant \citep{2011MNRAS.412...75S}.

@@

\begin{figure}
  \begin{center}
    \begin{tikzpicture}
      \draw [-angle 45,dashed,thick,dash pattern = on 7pt off 2pt] (0,0) -- (3.0,0);
      \draw [-angle 45,dashed,thick,dash pattern = on 7pt off 2pt] (0,0) -- (0,4.0);
      \draw [-angle 45,thick] (0,0) -- (2.0,4.0);
      \draw [-angle 45,thick,red] (0,0) -- (5.0,4.0);
      \draw (1.0,2.0) arc (63:90:2.2);
      \draw (2.0,1.6) arc (38.6:63:2.56);
      \draw (0.5,2.5) node {\large $\chi$};
      \draw (1.90,2.3) node {\large $\psi$};
      \draw [dashed,thick,dash pattern = on 7pt off 2pt] (0,0) -- (3.0,6.0);
      \draw (0.0,4.5) node {\large $\mathbf{L}_\mathrm{orb}$};
      \draw (3.5,0.0) node {\large $\mathbf{v}_\mathrm{orb}$};
      \draw (1.0,3.0) node {\large $\mathbf{v}_\mathrm{jet}$};
      \draw (5.3,4.2) node {\large \textcolor{red}{$\mathbf{\tilde v}_\mathrm{jet}$}};
    \end{tikzpicture}
  \end{center}
  \caption{The orbital velocity $\mathbf{v}_\mathrm{orb}$
    introduces the jet precession angle $\psi$.  Without the orbital
    motion, $\psi=0$, and the jet traces a cylindrical surface.  With
    the orbital motion, the net red velocity vector oscillates around
    the solid black vector and produces a conical surface with $\psi$
    as its half-opening angle.  Note that $v_\mathrm{orb}$ is
    exagerated here.  In practice, $v_\mathrm{orb}\ll
    v_\mathrm{jet}\sim c$.  The figure indicates that
    $\sin{\psi}\sim\tan{\psi}=v_\mathrm{orb}\cos{\chi}/v_\mathrm{jet}$.}
  \label{fig:jet}
\end{figure}

@@

\begin{figure*}
\begin{center}
  \begin{tabular}{cc}
    % jet.py (case 0) 
    \includegraphics*[scale=0.45]{{jet_i40_beta0.90_mdot1.00_full}.pdf} &
    % jet_image.py (case 1) 
    \includegraphics*[scale=0.45]{{jet_i40_beta0.90_mdot1.00_image_zoom1}.pdf}
  \end{tabular}
\end{center}
\caption{The left panel shows forward and backward jets for an
  equal-mass binary with total mass $M=10^{10}$~M$_\odot$.  (This
  figure shows the position of many jet particles observed at a fixed
  time on Earth.)  We have assumed $i=40^\circ$, $\beta=0.9$, and
  angular diameter distance $d = 100$~Mpc.  The jets are twisted
  because of the apparant superluminal motion.  Also evident is the
  opening of the jet and the radio jet chirp.  Right panel shows
  normalised brightness of the inner 30~mas region of the forward
  jet. (This panel shows the intensity of many particles observed at a
  fixed time on Earth.)}
\label{fig:jet_mdot1.0}
\end{figure*}

@@

As the binary evolves, the combination of increasing orbital speed and
the apparant superluminal motion effect results in a jet morphology
with an increasing opening angle and increasingly rapid periodic
variation, together with a twisted shape.
Figure~\ref{fig:jet_mdot1.0} shows these three features for an
equal-mass binary with total mass $M=10^{10}$~M$_\odot$.  For
concreteness, we have assumed $i=40^\circ$ and $\beta=0.9$.  The
angular diameter distance to the binary is assumed to be $d = 100$~Mpc
(which corresponds to redshift $z\sim 0.025$).  The forward jet
stretches on the sky while the backward jet is compressed due to the
apparent superluminal motion.  The right panel of
Figure~\ref{fig:jet_mdot1.0} zooms into the inner region of the
forward jet and shows that the binary creates wiggles in the jet
morphology of the order of a milli-arcsecond.  The duration of the
binary evolution captured by Figure~\ref{fig:jet_mdot1.0} is 100 yr.
Although the system can be rescaled to different masses, times, and
separation, the initial separation in Figure~\ref{fig:jet_mdot1.0} is
$\sim 10^{-3}$~pc, when the binary enters it GW-dominated phase.

The increasing orbital speed of the binary leads to an increase in the
jet's precession angular velocity.  This increases the helicity of the
jet such that the jet is wound up increasingly tightly closer to the
source.  This gives rise to a ``chirping'' morphology as seen in
Figure~\ref{fig:jet_mdot1.0}.  The chirp is easier to resolve in jets
that are closer to the line of sight, where superluminal motion can
magnify the forward jet.  A chirping black hole is an unambiguous
signature of a rapidly evolving binary BH system; the chirp is direct
reflection of the orbital velocity evolution of the binary. Together,
the three features of the jet morphology in
Figure~\ref{fig:jet_mdot1.0} are:
\begin{itemize}
\item Increase in the helicity of the jet towards the central source (``chirpyness''),
\item Widening of the opening angle towards the central source, and 
\item An apparant variation in the jet precession axis towards the central source (``twist''). 
\end{itemize}
These directly probe the dynamics of the binary in its GW phase.

@@

Observations of such a milliarc second scale geometry with the three
features described above can thus flag a BH binary nearing the end of
its merger.  We now consider the question of whether such morphologies
will be measurable in real observations.

As discussed in Section~\ref{sec:intro}, several helical jets have
been observed thanks to VLBI.  A resolution of few tens of microarc
seconds can be achieved using the Very Long Baseline Array (VLBA) with
sensitivity of up to 30 $\mu$Jy$/$beam, as demonstrated by the MOJAVE
survey \citep{2009AJ....138.1874L}.  VLA has angular resolution of few
tens of milliarcsecconds with a sensitivity of about 100
$\mu$Jy$/$beam.  The European VLBI network (EVN) which can resolve up
to 0.13 milliarc seconds and can detect surface brightness down to 19
$\mu$Jy$/$beam.  The Very Large Array (VLA) has an angular resolution
of few tens of milliarcsecconds with a sensitivity of about 100
$\mu$Jy$/$beam.  The future Square Kilometer Array (SKA) on its own
will have 2 milliarc second resolution at around 10 GHz, but SKA will
be an order of magnitude more sensitive than current instruments with
a sensitivity of about 50 nJy$/$beam.  Thus angular resolution of tens
to hundreds of microarc seconds and flux sensitivity of a few
$\mu$Jy$/$beam is achievable with current VLBI instruments.

The brighter part of the jet (intensity within factor of 10 from the
maximum) in Figure~\ref{fig:jet_mdot1.0} (right panel) has separations
of $\sim 1$ mas.  It should therefore be possible to resolve the jet
at an angular resolution of $\sim 2$ mas.  (In the next section, we
discuss an algorithm to fit the jet model to observational data.)  If
it is aligned closer to the line of sight, the forward jet is
magnified due to apparant superluminal motion.  This will make it
possible to resolve jets of objects that are farther away than the
example in Figure~\ref{fig:jet_mdot1.0}.  Binaries with higher masses
have large orbital periods that will result in a helix with wider
seperation between its cycles.  These objects would therefore be
easier to detect.  We have not modelled the absolute surface
brightness of jets, only relative variation in the jet brightness (as
shown in Figure~\ref{fig:jet_mdot1.0}), but due to the $(1+z)^{-4}$
dimming, it will be exceedingly difficult to detect such jets at
higher redshifts.

We note here that there is at least one object in the literature that
shows a jet morphology similar to Figure~\ref{fig:jet_mdot1.0}.  The
quasar 4C 18.68 described by \citet{1982ApJ...253L...1G} shows a
helical jet with an apparant opening angle that increases towards the
central source from 11$^\circ$ to 45$^\circ$ in just two cycles
\citet{1982ApJ...262..478G}.  The observing beamwidth of about 400 mas
makes it difficult to resolve a chirp in this source, but it is a
prototype of the morphology that we are discussing here.
\citet{2015MNRAS.453.1562G} have reported that 111 candidates show
periodic variability in brightness in the Cataline Real-time Transient
Survey, indicating presence of binary BHs.  An analysis of these
candidates suggests the presence of circumbinary gas at small orbital
distances \citep{2015MNRAS.453.1562G}.  This bodes well for the
likelihood of the presence of chirpy jets.  In low resolution wide
field survey, detecting modulations in the lightcurves would be an
easier diagnostic of a binary BH \citep{2011ApJ...743..136O,
  2011ApJ...734L..37K}.

\subsection{Deriving binary parameters from observations of chirping jets}

From observations of chirping jets, we can infer the dynamical
parameters of the binary BH by first fitting a purely geometric model
to the observed jet morphology.  To create a model of this kind, we
generalise the simple helical model considered by
\citet{2014MNRAS.445.1370K} and \citet{2001NewA....6...61N} to write
\begin{align}
  z &= x^{-\alpha} \frac{B}{2\pi} u \cos{u}, \\
  y &= x^{-\alpha} \frac{B}{2\pi} u \sin{u}, \\
  x &= \frac{A}{2\pi} u \left(\frac{u}{2\pi}\right)^\beta.
\end{align}
Here $u$ is the azimuthal angle around the jet axis, $x$ and $y$ axes
are in the plane of the sky, and the $z$ axis points to the observer.
The $x^{-\alpha}$ term in the first two equations captures the
widening opening angle of the jet, while the $(u/2\pi)^{\beta}$ term
in the last equation parameterises the chirp.  This model can be
fitted to an observed jet on the sky by using two rotation matrices,
$\mathcal{R}_\lambda$ and $\mathcal{R}_i$, which rotate the model in
the above equtions to match with the jet precession axis and the jet
symmetry axis.  These matrices are given by

\begin{flalign}
\mathrm{R}_{i} \!=\!
\begin{pmatrix}
\cos i & 0 & \sin i \\ 
0 & 1 & 0 \\ 
-\sin i & 0 & \cos i
\end{pmatrix}
\label{riota}
\end{flalign}
and
\begin{flalign}
\mathcal{R}_{\lambda}\!=\!
\begin{pmatrix}
\cos \lambda{_0} & -\sin \lambda{_0} & 0 \\ 
\sin \lambda{_0} & \cos \lambda{_0} & 0 \\ 
0 & 0 & 1%
\end{pmatrix}.
\label{rlambda}
\end{flalign}
Further, we can parameterise the ``twist'' in the jet by requiring
that $i$ and $\lambda$ vary with $x$.  Without loss of generality,
we take this variation in $i$ and parameterise it as
\begin{equation}
  i = \gamma x + i_0.
\end{equation}
We thus arrive at a model with 7 parameters ($\alpha$, $\beta$,
$\gamma$, $A$, $B$, $i_0$, $\lambda_0$), which can be fitted to any
observed jet, such as that in Figure~\ref{fig:jet_mdot1.0}.  Note that
setting $\alpha$ and $\beta$ equal to $0$ yields the usual helical jet
morphology, without the features discussed in this paper.  Thus, a
non-zero value of these two parameters distinguishes chirpy jets from
non-chirpy ones.

The best fit geometric parameters can then be used to derive the
dynamical parameters of the BH binary in the following way.  The
parameters $\alpha$ and $B$ quantify the widening of the opening angle
$\psi$.  As each cycle of the observed helical jet corresponds to a
period $P(t)$ of the binary, we can then write $d\psi/dP$, which
yields a constraint on the quantity $a^4/M^3$ as a function of $P$ via
Equations (\ref{eqn:period}), (\ref{eqn:shrink}) and
(\ref{eqn:widen}).  The parameters $A$ and $\beta$, which quantify the
chirp, can be used to directly infer the evolution of the binary
separation as $da/dP$, which also contraints the quantity $a^4/M^3$
via Equation~(\ref{eqn:shrink}). These two constraints can be used to
determine $a(P)$ and $M$ separately.  Finally, the parameters
$i_0$, $\lambda_0$, and $\gamma$ quantify the twist in the jet,
which can potentially be used to constrain the quantity
$\dot{\mathcal{M}}$ by using Equation~(\ref{eqn:v}).

@@

Using a model of jet precession in a binary BH, we have shown that the
jet produced by a member of a binary that is entering its GW-induced
phase of inspiral exhibits a distinct chirping morphology.  Jets in
these binaries have a biconical morphology.  After the binary enters
the GW phase, the opening angle of cone rapidly increases within a
separation of tens of milli-arcseconds at a distance of $\sim
100$~Mpc, depending on the angle that the jet makes relative to the
line of sight.  In addition, the periodicity of the jet's motion on
the conical surface increases with the orbital period of the binary in
this stage of its evolution.  Finally, due to the apparent
superluminal motion and the widening opening angle, the jet is twisted
into a peculiar morphology.  The jet geometry can be used to infer the
set of dynamical and geometrical parameters.

Milli-arcsecond features can be resolved using long baseline radio
interferometry \citep{2001ChJAA...1..236Q}.  At higher redshifts, jets
are fainter and the wiggles become more difficult to resolve.
However, it might be possible to resolve binaries with very unequal
mass ratios, eccentric orbits, or suitable inclinations.  The future
Square Kilometer Array (SKA) should make microarcsecond resolution
possible, which could increase the detection rate of these objects
\citep{2014arXiv1412.5971P}.

@@

We acknowledge helpful comments from the anonymous referee and useful
discussions with Martin Haehnelt, Enrico Ramirez-Ruiz and Alberto
Rorai.  This work was supported in part by NSF grant AST-1312034 (for
A.~L.) and by the FP7 ERC Advanced Grant Emergence-320596 (for G.~K.).

>>>>

\begin{abstract}
  We study a novel electromagnetic signature of supermassive black
  hole binaries whose inspiral starts being dominated by gravitational
  wave (GW) emission.  Recent simulations suggest that the binary's
  member BHs can continue to accrete gas from the circumbinary
  accretion disk in this phase of the binary's evolution, all the way
  until coalescence.  If one of the binary members produces a radio
  jet as a result of accretion, the jet precesses along a biconical
  surface due to the binary's orbital motion.  When the binary enters
  the GW phase of its evolution, the opening angle widens, the jet
  exhibits milliarcsecond scale wiggles, and the conical surface of
  jet precession is twisted due to apparent superluminal motion.  The
  rapidly increasing orbital velocity of the binary gives the jet an
  appearance of a ``chirp.''  This helical chirping morphology of the
  jet can be used to infer the binary parameters.  For binaries with
  mass $10^7$--$10^{10}$ M$_\odot$ at redshifts $z<0.5$, monitoring
  these features in current and archival data will place a lower limit
  on sources that could be detected by eLISA and Pulsar Timing Arrays.
  In the future, microarcsecond interferometry with the Square
  Kilometer Array will increase the potential usefulness of this
  technique.
\end{abstract}

@@

Binaries of supermassive black holes (BHs) arise naturally as a result
of mergers of galaxies in the context of hierarchical galaxy formation
\citep{2000MNRAS.311..576K, 2002MNRAS.336L..61H, 2008ApJ...676...33D,
  2012MNRAS.422.1306K}.  After a galaxy merger, infalling black holes
lose their angular momentum in three stages before coalescence
\citep{1980Natur.287..307B, 2005LRR.....8....8M, 2011ASL.....4..181C}.
In the first stage, the black holes sink to the center of the
gravitational potential of the merger remnant due to dynamical
friction and form a gravitationally bound binary.  The newly formed
binary then continues to lose its angular momentum and shrink by
scattering ambient gas and stars.  In the final stage of the binary's
evolution, emission of gravitational waves (GWs) becomes the
predominant mode of angular momentum loss, which results in the
coalescence of the black holes.

General relativistic (GR) numerical simulations can predict the
binary's evolution in its final stage up to coalescence
\citep{2005PhRvL..95l1101P, 2006PhRvL..96k1102B, 2006PhRvL..96k1101C}.
This has stimulated interest in predicting related electromagnetic
signals of black hole coalescence.  Effects considered in the
literature include afterglows due to infall of gas on the remnant
black hole \citep{2005ApJ...622L..93M}, electromagnetic variability in
the circumbinary disk due to shocks induced by the sudden mass loss of
the binary following GW emission at coalescence
\citep{2007APS..APR.S1010B, 2008ApJ...676L...5L}, flares from shocked
remnants of accretion disk around the recoiled remnant black hole
\citep{2007PhRvL..99d1103L, 2008ApJ...682..758S, 2009CQGra..26i4032H},
quasi-periodic variability in gas accretion in the early stages of
black hole merger \citep{2008ApJ...672...83M, 2009CQGra..26i4032H,
  2012MNRAS.427.2680K, 2013MNRAS.436.2997D, 2014ApJ...783..134F,
  2015ApJ...807..131S}, evolution in the profiles of broad emission
lines shortly before merger \citep{2013MNRAS.432.1468M},
characteristic evolution in thermal emission due to circumbinary
accretion \citep{2015MNRAS.446L..36F}, brightening of the circumbinary
accretion disk due to viscous dissipation of gravitational waves
\citep{2008PhRvL.101d1101K}, and prompt tidal disruption of ambient
stars by the recoiled remnant \citep{2011MNRAS.412...75S}.

@@

\begin{figure}
  \begin{center}
    \begin{tikzpicture}
      \draw [-angle 45,dashed,thick,dash pattern = on 7pt off 2pt] (0,0) -- (3.0,0);
      \draw [-angle 45,dashed,thick,dash pattern = on 7pt off 2pt] (0,0) -- (0,4.0);
      \draw [-angle 45,thick] (0,0) -- (2.0,4.0);
      \draw [-angle 45,thick,red] (0,0) -- (5.0,4.0);
      \draw (1.0,2.0) arc (63:90:2.2);
      \draw (2.0,1.6) arc (38.6:63:2.56);
      \draw (0.5,2.5) node {\large $\chi$};
      \draw (1.90,2.3) node {\large $\psi$};
      \draw [dashed,thick,dash pattern = on 7pt off 2pt] (0,0) -- (3.0,6.0);
      \draw (0.0,4.5) node {\large $\mathbf{L}_\mathrm{orb}$};
      \draw (3.5,0.0) node {\large $\mathbf{v}_\mathrm{orb}$};
      \draw (1.0,3.0) node {\large $\mathbf{v}_\mathrm{jet}$};
      \draw (5.3,4.2) node {\large \textcolor{red}{$\mathbf{\tilde v}_\mathrm{jet}$}};
    \end{tikzpicture}
  \end{center}
  \caption{The orbital velocity $\mathbf{v}_\mathrm{orb}$
    introduces the jet precession angle $\psi$.  Without the orbital
    motion, $\psi=0$, and the jet traces a cylindrical surface.  With
    the orbital motion, the net red velocity vector oscillates around
    the solid black vector and produces a conical surface with $\psi$
    as its half-opening angle.  Note that $v_\mathrm{orb}$ is
    exaggerated here.  In practice, $v_\mathrm{orb}\ll
    v_\mathrm{jet}\sim c$.  The figure indicates that
    $\sin{\psi}\sim\tan{\psi}=v_\mathrm{orb}\cos{\chi}/v_\mathrm{jet}$.}
  \label{fig:jet}
\end{figure}

@@

\begin{figure*}
\begin{center}
  \begin{tabular}{cc}
    % jet.py (case 0) 
    \includegraphics*[scale=0.45]{{jet_i40_beta0.90_mdot1.00_full}.pdf} &
    % jet_image.py (case 1) 
    \includegraphics*[scale=0.45]{{jet_i40_beta0.90_mdot1.00_image_zoom1}.pdf}
  \end{tabular}
\end{center}
\caption{The left panel shows forward and backward jets for an
  equal-mass binary with total mass $M=10^{10}$~M$_\odot$.  (This
  figure shows the position of many jet particles observed at a fixed
  time on Earth.)  We have assumed $i=40^\circ$, $\beta=0.9$, and
  angular diameter distance $d = 100$~Mpc.  The jets are twisted
  because of the apparent superluminal motion.  Also evident is the
  opening of the jet and the radio jet chirp.  Right panel shows
  normalised brightness of the inner 30~mas region of the forward
  jet. (This panel shows the intensity of many particles observed at a
  fixed time on Earth.)}
\label{fig:jet_mdot1.0}
\end{figure*}

@@

As the binary evolves, the combination of increasing orbital speed and
the apparent superluminal motion effect results in a jet morphology
with an increasing opening angle and increasingly rapid periodic
variation, together with a twisted shape.
Figure~\ref{fig:jet_mdot1.0} shows these three features for an
equal-mass binary with total mass $M=10^{10}$~M$_\odot$.  For
concreteness, we have assumed $i=40^\circ$ and $\beta=0.9$.  The
angular diameter distance to the binary is assumed to be $d = 100$~Mpc
(which corresponds to redshift $z\sim 0.025$).  The forward jet
stretches on the sky while the backward jet is compressed due to the
apparent superluminal motion.  The right panel of
Figure~\ref{fig:jet_mdot1.0} zooms into the inner region of the
forward jet and shows that the binary creates wiggles in the jet
morphology of the order of a milliarcsecond.  The duration of the
binary evolution captured by Figure~\ref{fig:jet_mdot1.0} is 100 yr.
Although the system can be rescaled to different masses, times, and
separation, the initial separation in Figure~\ref{fig:jet_mdot1.0} is
$\sim 10^{-3}$~pc, when the binary enters it GW-dominated phase.

The increasing orbital speed of the binary leads to an increase in the
jet's precession angular velocity.  This increases the helical winding
of the jet such that the jet is wound up increasingly tightly closer
to the source.  This gives rise to a ``chirping'' morphology as seen
in Figure~\ref{fig:jet_mdot1.0}.  The chirp is easier to resolve in
jets that are closer to the line of sight, where superluminal motion
can magnify the forward jet.  A chirping black hole is an unambiguous
signature of a rapidly evolving binary BH system; the chirp is direct
reflection of the orbital velocity evolution of the binary. Together,
the three features of the jet morphology in
Figure~\ref{fig:jet_mdot1.0} are:
\begin{itemize}
\item Increase in the helical winding of the jet towards the central source (``chirpyness''),
\item Widening of the opening angle towards the central source, and 
\item An apparant variation in the jet precession axis towards the central source (``twist''). 
\end{itemize}
These directly probe the dynamics of the binary in its GW phase.

@@

Observations of such a milliarcsecond scale geometry with the three
features described above can thus flag a BH binary nearing the end of
its merger.  We now consider the question of whether such morphologies
will be measurable in real observations.

As discussed in Section~\ref{sec:intro}, several helical jets have
been observed thanks to VLBI.  A resolution of few tens of microarc
seconds can be achieved using the Very Long Baseline Array (VLBA) with
sensitivity of up to 30 $\mu$Jy$/$beam, as demonstrated by the MOJAVE
survey \citep{2009AJ....138.1874L}.  VLA has angular resolution of few
tens of milliarcsecconds with a sensitivity of about 100
$\mu$Jy$/$beam.  The European VLBI network (EVN) which can resolve up
to 0.13 milliarcseconds and can detect surface brightness down to 19
$\mu$Jy$/$beam.  The Very Large Array (VLA) has an angular resolution
of few tens of milliarcsecconds with a sensitivity of about 100
$\mu$Jy$/$beam.  The future Square Kilometer Array (SKA) on its own
will have 2 milliarcsecond resolution at around 10 GHz, but SKA will
be an order of magnitude more sensitive than current instruments with
a sensitivity of about 50 nJy$/$beam.  Thus angular resolution of tens
to hundreds of microarc seconds and flux sensitivity of a few
$\mu$Jy$/$beam is achievable with current VLBI instruments.

The brighter part of the jet (intensity within factor of 10 from the
maximum) in Figure~\ref{fig:jet_mdot1.0} (right panel) has separations
of $\sim 1$ mas.  Figure~\ref{fig:smooth_jet} shows inner 100 square
mas of the jet in Figure~\ref{fig:jet_mdot1.0} smoothed at 200 $\mu$as
and 2 mas respectively.  We find that at 200 $\mu$as resolution the
chirp is clearly resolved.  At a lower resolution of 2 mas, the chirp
is no longer resolved but widening of the jet opening angle can
possibly be inferred.  (In the next section, we discuss an algorithm
to fit the jet model to observational data.)  Note that other
geometric effects can influence these conclusions.  If it is aligned
closer to the line of sight, the forward jet is magnified due to
apparant superluminal motion.  This will make it possible to resolve
jets of objects that are farther away than the example in
Figure~\ref{fig:jet_mdot1.0}.  Binaries with higher masses have large
orbital periods that will result in a helix with wider separation
between its cycles.  These objects would therefore be easier to
detect.  We have not modeled the absolute surface brightness of jets,
only relative variation in the jet brightness (as shown in
Figure~\ref{fig:jet_mdot1.0}), but due to the $(1+z)^{-4}$ dimming, it
will be exceedingly difficult to detect such jets at higher redshifts.

We note here that there is at least one object in the literature that
shows a jet morphology similar to Figure~\ref{fig:jet_mdot1.0}.  The
quasar 4C 18.68 described by \citet{1982ApJ...253L...1G} shows a
helical jet with an apparant opening angle that increases towards the
central source from 11$^\circ$ to 45$^\circ$ in just two cycles
\citep{1982ApJ...262..478G}.  The observing beamwidth of about 400 mas
makes it difficult to resolve a chirp in this source, but it is a
prototype of the morphology that we are discussing here.
\citet{2015MNRAS.453.1562G} have reported that 111 candidates show
periodic variability in brightness in the Catalina Real-time Transient
Survey, indicating presence of binary BHs.  An analysis of these
candidates suggests the presence of circumbinary gas at small orbital
distances \citep{2015MNRAS.453.1562G}.  This bodes well for the
likelihood of the presence of chirpy jets.  In low resolution wide
field surveys, detecting modulations in the light curves would be an
easier diagnostic of a binary BH \citep{2011ApJ...743..136O,
  2011ApJ...734L..37K}.

\begin{figure*}
\begin{center}
  \begin{tabular}{cc}
    % model.py 
    \includegraphics*[width=\columnwidth]{jet_200muas.pdf} &
    % model.py 
    \includegraphics*[width=\columnwidth]{jet_2mas.pdf}
  \end{tabular}
\end{center}
\caption{Central 100 square mas of the jet in
  Figure~\ref{fig:jet_mdot1.0} smoothed at 200 $\mu$as (left panel)
  and 2 mas (right panel) respectively.  The intensity is normalised
  in arbitrary units.}
\label{fig:smooth_jet}
\end{figure*}

\subsection{Deriving binary parameters from observations of chirping jets}
\label{sec:geometric_model}

From observations of chirping jets, we can infer the dynamical
parameters of the binary BH by first fitting a purely geometric model
to the observed jet morphology.  To create a model of this kind, we
generalise the simple helical model considered by
\citet{2014MNRAS.445.1370K} and \citet{2001NewA....6...61N} to write
\begin{align}
  z &= x^{-\alpha} \frac{B}{2\pi} u \cos{u}, \label{eqn:z}\\
  y &= x^{-\alpha} \frac{B}{2\pi} u \sin{u}, \\
  x &= \frac{A}{2\pi} u \left(\frac{u}{2\pi}\right)^\beta.
\end{align}
Here $u$ is the azimuthal angle around the jet axis, $x$ and $y$ axes
are in the plane of the sky, and the $z$ axis points to the observer.
The $x^{-\alpha}$ term in the first two equations captures the
widening opening angle of the jet, while the $(u/2\pi)^{\beta}$ term
in the last equation parameterises the chirp.  This model can be
fitted to an observed jet on the sky by using two rotation matrices,
$\mathcal{R}_\lambda$ and $\mathcal{R}_i$, which rotate the model in
the above equations to match with the jet precession axis and the jet
symmetry axis.  These matrices are given by

\begin{flalign}
\mathrm{R}_{i} \!=\!
\begin{pmatrix}
\cos i & 0 & \sin i \\ 
0 & 1 & 0 \\ 
-\sin i & 0 & \cos i
\end{pmatrix}
\label{riota}
\end{flalign}
and
\begin{flalign}
\mathcal{R}_{\lambda}\!=\!
\begin{pmatrix}
\cos \lambda{_0} & -\sin \lambda{_0} & 0 \\ 
\sin \lambda{_0} & \cos \lambda{_0} & 0 \\ 
0 & 0 & 1%
\end{pmatrix}.
\label{rlambda}
\end{flalign}
Further, we can parameterise the ``twist'' in the jet by requiring
that $i$ and $\lambda$ vary with $x$.  Without loss of generality,
we take this variation in $i$ and parameterise it as
\begin{equation}
  i = \gamma x + i_0.
  \label{eqn:i}
\end{equation}
We thus arrive at a model with 7 parameters ($\alpha$, $\beta$,
$\gamma$, $A$, $B$, $i_0$, $\lambda_0$), which can be fitted to any
observed jet, such as that in Figure~\ref{fig:jet_mdot1.0}.  Note that
setting $\alpha$ and $\beta$ equal to $0$ yields the usual helical jet
morphology, without the features discussed in this paper.  Thus, a
non-zero value of these two parameters distinguishes chirpy jets from
non-chirpy ones.

\begin{figure*}
\begin{center}
  \includegraphics*[width=\textwidth]{triangle.pdf}
\end{center}
\caption{One and two dimensional projections of posterior probability
  distributions of the 7-parameter geometric model described in
  Section~\ref{sec:geometric_model} to mock data.  Wide, uniform
  priors were chosen for all parameters.  The two-dimensional
  projections also show that there are no parameter degeneracies in
  the geometric model.  As described in the text, these parameter
  values can then be used to derive various physical parameters
  describing the black hole binary.}
\label{fig:triangle}
\end{figure*}

Figure \ref{fig:triangle} shows how the geometric model of this
section works in practice, by fitting it to mock data using Markov
Chain Monte Carlo (MCMC).  We used the \texttt{emcee} package for this
purpose \citep{2013PASP..125..306F}.  Mock data are created by using
Equations~(\ref{eqn:z})--(\ref{eqn:i}) and then smoothing the
resultant jet morphology by $\sim$ 2 mas.  Figure \ref{fig:triangle}
shows posterior probability distributions of the parameters of our
7-parameter model when it is fit to these mock data.  Wide, uniform
priors were chosen for all parameters.  The posteriors demonstrate
that there is no degeneracy in the geometric model parameters.  In
fact the narrow ranges of values for which the posteriors are non-zero
for each parameter show that the model fit the mock data very well.

The best fit geometric parameters can then be used to derive the
dynamical parameters of the BH binary in the following way.  The
parameters $\alpha$ and $B$ quantify the widening of the opening angle
$\psi$.  As each cycle of the observed helical jet corresponds to a
period $P(t)$ of the binary, we can then write $d\psi/dP$, which
yields a constraint on the quantity $a^4/M^3$ as a function of $P$ via
Equations (\ref{eqn:period}), (\ref{eqn:shrink}) and
(\ref{eqn:widen}).  The parameters $A$ and $\beta$, which quantify the
chirp, can be used to directly infer the evolution of the binary
separation as $da/dP$, which also constraints the quantity $a^4/M^3$
via Equation~(\ref{eqn:shrink}). These two constraints can be used to
determine $a(P)$ and $M$ separately.  Finally, the parameters
$i_0$, $\lambda_0$, and $\gamma$ quantify the twist in the jet,
which can potentially be used to constrain the quantity
$\dot{\mathcal{M}}$ by using Equation~(\ref{eqn:v}).

@@

Using a model of jet precession in a binary BH, we have shown that the
jet produced by a member of a binary that is entering its GW-induced
phase of inspiral exhibits a distinct chirping morphology.  Jets in
these binaries have a biconical morphology.  After the binary enters
the GW phase, the opening angle of cone rapidly increases within a
separation of tens of milliarcseconds at a distance of $\sim 100$~Mpc,
depending on the angle that the jet makes relative to the line of
sight.  In addition, the periodicity of the jet's motion on the
conical surface increases with the orbital period of the binary in
this stage of its evolution.  Finally, due to the apparent
superluminal motion and the widening opening angle, the jet is twisted
into a peculiar morphology.  The jet geometry can be used to infer the
set of dynamical and geometrical parameters.

Milliarcsecond features can be resolved using long baseline radio
interferometry \citep{2001ChJAA...1..236Q}.  At higher redshifts, jets
are fainter and the wiggles become more difficult to resolve.
However, it might be possible to resolve binaries with very unequal
mass ratios, eccentric orbits, or suitable inclinations.  The future
Square Kilometer Array (SKA) should make microarcsecond resolution
possible, which could increase the detection rate of these objects
\citep{2015aska.confE.143P}.

@@

We acknowledge helpful comments from the anonymous referee and useful
discussions with Martin Haehnelt, Enrico Ramirez-Ruiz, and Alberto
Rorai.  This work was supported in part by NSF grant AST-1312034 (for
A.~L.) and by the FP7 ERC Advanced Grant Emergence-320596 (for G.~K.).
